---
title: "VOC replacement forecast model"
output: 
  html_notebook:
    code_folding: hide
---

## Background
ECDC uses a logistic growth model to estimate the time until an emergent SARS-CoV-2 variant becomes the dominant variant (>50%) in circulation (see https://www.ecdc.europa.eu/sites/default/files/documents/Epi-update-BQ1.pdf). The model is adapted from models developed by Althaus et al. 2021 (https://www.medrxiv.org/content/10.1101/2021.06.10.21258468v1.full.pdf) and Chen et al. 2021 (https://www.sciencedirect.com/science/article/pii/S1755436521000335?via%3Dihub). 

The model estimates the daily proportion of cases due to the new variant as a function of two parameters: mu, the ratio of variant to other cases at time 0, and rho, the daily growth advantage of the new variant relative to existing variants. Here, mu is estimated from the proportion of sequenced isolates belonging to the new variant at time 0. rho is the growth in proportions relative to all other circulating lineages (estimates for Canada are from NLM, international estimates from CoV-Spectrum, https://cov-spectrum.org/explore/Europe/AllSamples/Past6M)

[Assumptions, limitations]

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 9)
```

```{r}
## load libraries
library(tidyverse)
library(here)
```

```{r}
## define function to estimate daily proportion of variant
estimate_pt <- function(t, mu, rho) {
  
  x <- 1 / (1 + mu * exp(1)^-(rho * t))
  
}

```

```{r}
## define parameter values
p_voi <- 0.01
mu <- (1 - p_voi) / p_voi
rho <- 0.11
start_date <- as.Date("2022-10-08")
```

```{r}
dat <- tibble(
  t = c(0:120),
  date = start_date + t,
  pt = estimate_pt(t, mu, rho)
)

```

```{r}
## plot
dat %>%
  ggplot(aes(x = date, y = pt)) +
  geom_line() +
  scale_y_continuous(breaks = seq(0, 1, 0.2))
```
```{r}
## model incorporating uncertainty in mu and rho parameters

## function
simulate_pt <- function(d, ...) {
  
  ## sample rho
  rho <- rnorm(1, rho_mean, rho_sd)
  
  ## sample logit(p_voi)
  logit_p <- rnorm(1, p_voi_mean, p_voi_sd)
  p_voi <- boot::inv.logit(logit_p) # calculate initial proportion
  mu <- (1 - p_voi) / p_voi
  
  ## calculate pt
  d <- d %>%
    mutate(
      pt = 1 / (1 + mu * exp(1)^-(rho * d$t))
    )
  
  d
  #x <- rho * d$t
  #return(x)
  #print(rho)
  #print(mu)
  #print(x)
}

## rho is a random draw from ~N(rho_mean, rho_sd)
rho_mean <- 0.11
rho_sd <- 0.02

## logit(p_voi) is a random draw from ~N(p_voi_mean, p_voi_sd)
p_voi_mean <- -4.6
p_voi_sd <- 1.5

## number of random draws
N <- 100

## set time period in days
t_end <- 120
start_date <- as.Date("2022-10-08")

## set random seed
set.seed(20230330)

```

```{r}
## create nested dataframe for simulation
dat2 <- tibble(
  t = rep(c(1:t_end), N),
  run = rep(c(1:N), each = t_end)
)

dat2 <- dat2 %>%
  group_by(run) %>%
  group_modify(simulate_pt) %>%
  ungroup()
```

```{r}
dat2_summary <- dat2 %>%
  group_by(t) %>%
  summarise(
    median = median(pt),
    p5 = quantile(pt, probs = 0.05),
    p95 = quantile(pt, probs = 0.95),
    p25 = quantile(pt, probs = 0.25),
    p75 = quantile(pt, probs = 0.75)
  ) %>%
  mutate(
    date = start_date + t
  )
```

```{r}
## plot
dat2_summary %>%
  ggplot() + 
  geom_ribbon(aes(x = date, ymin = p5, ymax = p95), fill = "gray80", alpha = 0.5) +
  geom_ribbon(aes(x = date, ymin = p25, ymax = p75), fill = "gray50", alpha = 0.5) +
  geom_line(aes(x = date, y = median), col = "black") +
  scale_y_continuous(breaks = seq(0, 1, 0.2), labels = scales::percent) + 
  scale_x_date(date_labels = "%b-%y") +
  labs(
    y = "Percentage of variant"
  )
```

