---
title: "voc_forecasts_generic"
format: html
editor: visual
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 9)
```

## Estimating variant-specific *R~t~* from COVID-19 daily case and variant proportion data

This script runs a series of functions to load data on daily COVID-19 cases and variant proportions, smooth the data, apply variant proportions to overall cases to estimate daily cases due to a variant, and estimate *R~t~*. *R~t~* estimates can also be used to generate short-term forecasts using the EpiNow2 package.

Note that these estimates do not account for uncertainty resulting from the number of sequences available, and assume fixed values for the serial interval mean and standard deviation.

The functions are stored in file functions.R

```{r}
## load functions
source(here::here("code/functions", "functions.R"))
```

The functions are:

-   get_case_data - download and process data from opencovid.ca. This function takes a single argument corresponding to the date up to which data should be downloaded and creates a variable with a 7-day moving average of daily case counts

-   get_voc_data - load and process variant-specific data downloaded from CoV-Spectrum. This function takes a single argument corresponding to the name of the CSV file containing the variant proportions data and creates variables for daily smoothed variant proportions (and 95% CI) using a LOESS smoother with span = 0.1

-   estimate_variant_cases - apply variant proportions to smoothed daily case data shifted by a specified number of days to align sequence and case data (accounting for reporting delays). This function takes 3 arguments corresponding to the dataframe containing case data, the dataframe containing variant proportions data, and an integer defining the shift in days to be used for aligning case and variant data, to account for reporting delays in the case data

-   get_R\_estimates - estimate R(t) from smoothed estimates of daily variant cases. This function takes 3 arguments corresponding to the list containing estimated variant case data, and assumed values for the mean and standard deviation of the serial interval (a gamma distribution is assumed for the serial interval by default)

Download case data and store in a dataframe:

```{r}
## download and process COVID-19 case data
case_data <- get_case_data("2023-05-02")
```

Load file containing variant proportions data:

```{r}
voc_data <- get_voc_data("VariantTimeDistributionPlot.xbb.1.9.1.csv")
```

Align case and variant proportion time series and estimate daily variant cases. Cases are estimated from the date the variant was first detected. The resulting data and plot are stored in a list:

```{r}
voc_cases <- estimate_variant_cases(case_data, voc_data, 7)

names(voc_cases)
```

Plot daily variant cases

```{r}
print(voc_cases$plot)
```

Estimate *R~t~* assuming a mean of 3.3 and standard deviation of 3.5 for the serial interval (obtained from estimates of early Omicron variants by Abbott et al. https://www.medrxiv.org/content/10.1101/2022.01.08.22268920v1.full.pdf)

*R~t~* estimates are derived using the EpiEstim package, assuming a 7-day sliding window for *R~t~*:

```{r}
variant_r <- get_R_estimates(voc_cases, 3.3, 3.5)

names(variant_r)
```

Plot results

```{r}
print(variant_r$plot)
```
