---
title: "VOC replacement forecast model"
output:
  html_notebook:
    code_folding: hide
---

## Background
ECDC uses a logistic growth model to estimate the time until an emergent SARS-CoV-2 variant becomes the dominant variant (>50%) in circulation (see https://www.ecdc.europa.eu/sites/default/files/documents/Epi-update-BQ1.pdf). The model is adapted from models developed by Althaus et al. 2021 (https://www.medrxiv.org/content/10.1101/2021.06.10.21258468v1.full.pdf) and Chen et al. 2021 (https://www.sciencedirect.com/science/article/pii/S1755436521000335?via%3Dihub). 

The model estimates the daily proportion of cases due to the new variant as a function of two parameters: mu, the ratio of variant to other cases at time 0, and rho, the daily growth advantage of the new variant relative to existing variants. Here, mu is estimated from the proportion of sequenced isolates belonging to the new variant at time 0. rho is the growth in proportions relative to all other circulating lineages (estimates for Canada are from NLM, international estimates from CoV-Spectrum, https://cov-spectrum.org/explore/Europe/AllSamples/Past6M)

[Assumptions, limitations]

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 9)
```

```{r}
## load libraries
library(tidyverse)
library(here)
```

```{r}
## define function to estimate daily proportion of variant
estimate_pt <- function(t, mu, rho) {
  
  x <- 1 / (1 + mu * exp(1)^-(rho * t))
  
}

```

```{r}
## define parameter values
p_voi <- 0.01
mu <- (1 - p_voi) / p_voi
rho <- 0.11
start_date <- as.Date("2022-10-08")
```

```{r}
dat <- tibble(
  t = c(0:120),
  date = start_date + t,
  pt = estimate_pt(t, mu, rho)
)

```

```{r}
## plot
dat %>%
  ggplot(aes(x = date, y = pt)) +
  geom_line() +
  scale_y_continuous(breaks = seq(0, 1, 0.2))
```
```{r}
## model incorporating uncertainty in mu and rho parameters

## function
simulate_pt <- function(d, ...) {
  
  ## sample rho
  rho <- rnorm(1, rho_mean, rho_sd)
  
  ## sample logit(p_voi)
  logit_p <- rnorm(1, p_voi_mean, p_voi_sd)
  p_voi <- boot::inv.logit(logit_p) # calculate initial proportion
  mu <- (1 - p_voi) / p_voi
  
  ## calculate pt
  d <- d %>%
    mutate(
      pt = 1 / (1 + mu * exp(1)^-(rho * d$t))
    )
  
  d
  #x <- rho * d$t
  #return(x)
  #print(rho)
  #print(mu)
  #print(x)
}

## rho is a random draw from ~N(rho_mean, rho_sd)
rho_mean <- 0.11
rho_sd <- 0.02

## logit(p_voi) is a random draw from ~N(p_voi_mean, p_voi_sd)
p_voi_mean <- -4.6
p_voi_sd <- 1.5

## number of random draws
N <- 100

## set time period in days
t_end <- 120
start_date <- as.Date("2022-10-08")

## set random seed
set.seed(20230330)

```

```{r}
## create nested dataframe for simulation
dat2 <- tibble(
  t = rep(c(1:t_end), N),
  run = rep(c(1:N), each = t_end)
)

dat2 <- dat2 %>%
  group_by(run) %>%
  group_modify(simulate_pt) %>%
  ungroup()
```

```{r}
dat2_summary <- dat2 %>%
  group_by(t) %>%
  summarise(
    median = median(pt),
    p5 = quantile(pt, probs = 0.05),
    p95 = quantile(pt, probs = 0.95),
    p25 = quantile(pt, probs = 0.25),
    p75 = quantile(pt, probs = 0.75)
  ) %>%
  mutate(
    date = start_date + t
  )
```

```{r}
## plot
dat2_summary %>%
  ggplot() + 
  geom_ribbon(aes(x = date, ymin = p5, ymax = p95), fill = "gray80", alpha = 0.5) +
  geom_ribbon(aes(x = date, ymin = p25, ymax = p75), fill = "gray50", alpha = 0.5) +
  geom_line(aes(x = date, y = median), col = "black") +
  scale_y_continuous(breaks = seq(0, 1, 0.2), labels = scales::percent) + 
  scale_x_date(date_labels = "%b-%y") +
  labs(
    y = "Percentage of variant",
    x = "",
    title = "Estimated daily percentage of BQ.1/BQ.1.1 cases among all SARS-CoV-2 cases in the EU/EEA"
  )
```
```{r}
## NOTE: data from CoV-SPECTRUM are cumulative not week-specific
## load XBB1.5 historical data from CoV-SPECTRUM
# xbb <- read.csv(here("data", "cov-spectrum_xbb15.csv"), stringsAsFactors = FALSE)
# xbb <- xbb %>%
#   mutate(date = as.Date(date))

## Download weekly XBB.1.5 proportions data from covariants.org (https://covariants.org/per-country)
## data are available here: https://raw.githubusercontent.com/hodcroftlab/covariants/master/cluster_tables/EUClusters_data.json
## note that these are based on Nextstrain clade assignment (23A Omicron)

library(jsonlite)

url <- "https://raw.githubusercontent.com/hodcroftlab/covariants/master/cluster_tables/EUClusters_data.json"
json_data <- fromJSON(url, simplifyDataFrame = TRUE)
us <- as_tibble(json_data$countries$US) %>% mutate(country = "United States") #%>% select(country, week, total_sequences, contains("23A"))
ca <- as_tibble(json_data$countries$Canada) %>% mutate(country = "Canada") #%>% select(country, week, total_sequences, contains("23A"))
sg <- as_tibble(json_data$countries$Singapore) %>% mutate(country = "Singapore") #%>% select(country, week, total_sequences, contains("23A"))
fr <- as_tibble(json_data$countries$France) %>% mutate(country = "France") #%>% select(country, week, total_sequences, contains("23A"))
de <- as_tibble(json_data$countries$Germany) %>% mutate(country = "Germany") #%>% select(country, week, total_sequences, contains("23A"))
uk <- as_tibble(json_data$countries$`United Kingdom`) %>% mutate(country = "United Kingdom") #%>% select(country, week, total_sequences, contains("23A"))

xbb <- bind_rows(us, ca, sg, fr, de, uk)

xbb <- xbb %>%
  select(country, week, total_sequences, contains("23A")) 

names(xbb) <- c("country", "date", "total_sequences", "xbb_sequences")

xbb <- xbb %>%
  mutate(
    xbb_prop = xbb_sequences / total_sequences,
    date = as.Date(date)
  )

## calculate binomial 95% CIs
conf_intervals <- as_tibble(Hmisc::binconf(xbb$xbb_sequences, xbb$total_sequences, alpha = 0.05))
xbb <- bind_cols(xbb, conf_intervals %>% select(-PointEst))

```

```{r,fig.width = 12, fig.height = 6}
## plot proportion XBB1.5 over time
xbb %>%
  ggplot(aes(x = date, y = xbb_prop, group = country, col = country)) +
  geom_line() +
  scale_x_date(date_labels = "%b-%y")
```

```{r}
## run model with country-specific parameters based on proportion of XBB1.5 on a specific month
start_month <- lubridate::ym("2022-12")

## extract country-specific proportions on start date
xbb <- xbb %>%
  mutate(
    ym = lubridate::ym(str_sub(as.character(date), 1, 7))
  )
         
start_df <- xbb %>%
  filter(ym == start_month) %>%
  filter(date == min(date)) %>%
  arrange(country)

```

```{r}
countries <- start_df %>% select(country) %>% pull() # vector of country names
start_props <- start_df %>% select(xbb_prop) %>% pull() # vector of country-specific variant proportions
start_date <- min(start_df$date) # start date of analysis
t_end <- 120 # length of estimation period in days
rho <- 0.08 # assumed growth advantage for new variant at the start of the analysis period
rho_min <- rho * 0.75 # if growth advantage is 25% lower or 50% higher
rho_max <- rho * 1.25

estimate_pt_country <- function(input_countries, p) {
  
  mu <- (1 - p) / p
  
  ## calculate date when proportion reaches 50%
  t50 <- -1 * rho^-1 * (log(1-0.5) - log(mu * 0.5))
  
  d <- tibble(
    country = rep(input_countries, t_end+1),
    t = c(0:t_end),
    date = start_date + t
  )
  
  d <- d %>%
    mutate(
      pt = 1 / (1 + mu * exp(1)^-(rho * t)),
      pt_min = 1 / (1 + mu * exp(1)^-(rho_min * t)),
      pt_max = 1 / (1 + mu * exp(1)^-(rho_max * t)),
      t50 = start_date + t50
    )

}

country_list <- map2(countries, start_props, estimate_pt_country)
xbb_predict <- country_list %>% bind_rows()

## merge with observed data
xbb_predict <- xbb_predict %>%
  left_join(xbb, by = c("country", "date"))

## extract mid-points for plotting
xbb_midpoint <- xbb_predict %>%
  group_by(country) %>%
  summarise(t50 = min(t50))
```

```{r, fig.width=16, fig.height=8}
## plot
xbb_predict %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = pt)) +
  geom_ribbon(aes(ymin = pt_min, ymax = pt_max), fill = "gray80", alpha = 0.5) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), size = 0.5, col = "gray50") +
  geom_point(aes(y = xbb_prop)) +
  geom_hline(aes(yintercept = 0.5), col = "black", linetype = "dotted") +
  geom_vline(data = xbb_midpoint, aes(xintercept = t50), col = "black", linetype = "dotted") +
  facet_wrap(~country) +
  scale_x_date(date_labels = "%b-%y") +
  labs(
    x = "",
    y = "Proportion XBB.1.5",
    caption = "Shaded area represents uncertainty bounds based on growth advantage being 25% lower or higher"
  ) 
```

```{r}
## Canada province/territory-specific data
## from: https://virusseq-dataportal.ca/explorer
## data downloaded on 2023-04-04 for the period from 2023-10-01 onwards
ca_variants_by_pt <- read_tsv(here("data", "virusseq-metadata-export-2023-04-04T17_05_45.405Z.tsv"))

## count XBB1.5 and total sequences by PT and sample collection date
xbb_by_pt <- ca_variants_by_pt %>%
  mutate(
    xbb15 = if_else(`lineage name` == "XBB.1.5", 1, 0)
  ) %>%
  group_by(study_id, `sample collection date`) %>%
  summarise(xbb15 = sum(xbb15), total = n())
```

