---
title: "VOC replacement forecast model"
output:
  html_notebook:
    code_folding: hide
---

## Background
ECDC uses a logistic growth model to estimate the time until an emergent SARS-CoV-2 variant becomes the dominant variant (>50%) in circulation (see https://www.ecdc.europa.eu/sites/default/files/documents/Epi-update-BQ1.pdf). The model is adapted from models developed by Althaus et al. 2021 (https://www.medrxiv.org/content/10.1101/2021.06.10.21258468v1.full.pdf) and Chen et al. 2021 (https://www.sciencedirect.com/science/article/pii/S1755436521000335?via%3Dihub). 

The model estimates the daily proportion of cases due to the new variant as a function of two parameters: mu, the ratio of variant to other cases at time 0, and rho, the daily growth advantage of the new variant relative to existing variants. Here, mu is estimated from the proportion of sequenced isolates belonging to the new variant at time 0. rho is the growth in proportions relative to all other circulating lineages (estimates for Canada are from NLM, international estimates from CoV-Spectrum, https://cov-spectrum.org/explore/Europe/AllSamples/Past6M)

[Assumptions, limitations]

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 9)
```

```{r}
## load libraries
library(tidyverse)
library(here)
```

```{r}
## define function to estimate daily proportion of variant
estimate_pt <- function(t, mu, rho) {
  
  x <- 1 / (1 + mu * exp(1)^-(rho * t))
  
}

```

```{r}
## define parameter values
p_voi <- 0.01
mu <- (1 - p_voi) / p_voi
rho <- 0.11
start_date <- as.Date("2022-10-08")
```

```{r}
dat <- tibble(
  t = c(0:120),
  date = start_date + t,
  pt = estimate_pt(t, mu, rho)
)

```

```{r}
## plot
dat %>%
  ggplot(aes(x = date, y = pt)) +
  geom_line() +
  scale_y_continuous(breaks = seq(0, 1, 0.2))
```
```{r}
## model incorporating uncertainty in mu and rho parameters

## function
simulate_pt <- function(d, ...) {
  
  ## sample rho
  rho <- rnorm(1, rho_mean, rho_sd)
  
  ## sample logit(p_voi)
  logit_p <- rnorm(1, p_voi_mean, p_voi_sd)
  p_voi <- boot::inv.logit(logit_p) # calculate initial proportion
  mu <- (1 - p_voi) / p_voi
  
  ## calculate pt
  d <- d %>%
    mutate(
      pt = 1 / (1 + mu * exp(1)^-(rho * d$t))
    )
  
  d
  #x <- rho * d$t
  #return(x)
  #print(rho)
  #print(mu)
  #print(x)
}

## rho is a random draw from ~N(rho_mean, rho_sd)
rho_mean <- 0.11
rho_sd <- 0.02

## logit(p_voi) is a random draw from ~N(p_voi_mean, p_voi_sd)
p_voi_mean <- -4.6
p_voi_sd <- 1.5

## number of random draws
N <- 100

## set time period in days
t_end <- 120
start_date <- as.Date("2022-10-08")

## set random seed
set.seed(20230330)

```

```{r}
## create nested dataframe for simulation
dat2 <- tibble(
  t = rep(c(1:t_end), N),
  run = rep(c(1:N), each = t_end)
)

dat2 <- dat2 %>%
  group_by(run) %>%
  group_modify(simulate_pt) %>%
  ungroup()
```

```{r}
dat2_summary <- dat2 %>%
  group_by(t) %>%
  summarise(
    median = median(pt),
    p5 = quantile(pt, probs = 0.05),
    p95 = quantile(pt, probs = 0.95),
    p25 = quantile(pt, probs = 0.25),
    p75 = quantile(pt, probs = 0.75)
  ) %>%
  mutate(
    date = start_date + t
  )
```

```{r}
## plot
dat2_summary %>%
  ggplot() + 
  geom_ribbon(aes(x = date, ymin = p5, ymax = p95), fill = "gray80", alpha = 0.5) +
  geom_ribbon(aes(x = date, ymin = p25, ymax = p75), fill = "gray50", alpha = 0.5) +
  geom_line(aes(x = date, y = median), col = "black") +
  scale_y_continuous(breaks = seq(0, 1, 0.2), labels = scales::percent) + 
  scale_x_date(date_labels = "%b-%y") +
  labs(
    y = "Percentage of variant",
    x = "",
    title = "Estimated daily percentage of BQ.1/BQ.1.1 cases among all SARS-CoV-2 cases in the EU/EEA"
  )
```
```{r}
## NOTE: data from CoV-SPECTRUM are cumulative not week-specific
## load XBB1.5 historical data from CoV-SPECTRUM
#xbb <- read.csv(here("data", "cov-spectrum_xbb15.csv"), stringsAsFactors = FALSE)
# xbb <- xbb %>%
#   mutate(date = as.Date(date))

## Download weekly XBB.1.5 proportions data from covariants.org (https://covariants.org/per-country)
## data are available here: https://raw.githubusercontent.com/hodcroftlab/covariants/master/cluster_tables/EUClusters_data.json
## note that these are based on Nextstrain clade assignment (23A Omicron)

library(jsonlite)

url <- "https://raw.githubusercontent.com/hodcroftlab/covariants/master/cluster_tables/EUClusters_data.json"
json_data <- fromJSON(url, simplifyDataFrame = TRUE)
us <- as_tibble(json_data$countries$US) %>% mutate(country = "United States") #%>% select(country, week, total_sequences, contains("23A"))
ca <- as_tibble(json_data$countries$Canada) %>% mutate(country = "Canada") #%>% select(country, week, total_sequences, contains("23A"))
sg <- as_tibble(json_data$countries$Singapore) %>% mutate(country = "Singapore") #%>% select(country, week, total_sequences, contains("23A"))
fr <- as_tibble(json_data$countries$France) %>% mutate(country = "France") #%>% select(country, week, total_sequences, contains("23A"))
de <- as_tibble(json_data$countries$Germany) %>% mutate(country = "Germany") #%>% select(country, week, total_sequences, contains("23A"))
uk <- as_tibble(json_data$countries$`United Kingdom`) %>% mutate(country = "United Kingdom") #%>% select(country, week, total_sequences, contains("23A"))

xbb <- bind_rows(us, ca, sg, fr, de, uk)

xbb <- xbb %>%
  select(country, week, total_sequences, contains("23A")) 

names(xbb) <- c("country", "date", "total_sequences", "xbb_sequences")

xbb <- xbb %>%
  mutate(
    xbb_prop = xbb_sequences / total_sequences,
    date = as.Date(date)
  )

## calculate binomial 95% CIs
conf_intervals <- as_tibble(Hmisc::binconf(xbb$xbb_sequences, xbb$total_sequences, alpha = 0.05))
xbb <- bind_cols(xbb, conf_intervals %>% select(-PointEst))

```

```{r,fig.width = 12, fig.height = 6}
## plot proportion XBB1.5 over time
xbb %>%
  ggplot(aes(x = date, y = xbb_prop, group = country, col = country)) +
  geom_line() +
  scale_x_date(date_labels = "%b-%y")
```
```{r}
## exclude weeks with no XBB.1.5 detection
xbb <- xbb %>%
  filter(xbb_sequences > 0)

## logistic regression to estimate growth advantage by country based on first 2 months of data (5 2-week periods)
xbb <- xbb %>%
  group_by(country) %>%
  mutate(
    t = row_number()
  )
  

growth_estimates <- xbb %>%
  filter(t <= 5) %>%
  group_modify(
    # Use `tidy`, `glance` or `augment` to extract different information from the fitted models.
    ~ broom::tidy(glm(xbb_prop ~ t, data = ., weights = total_sequences, family = binomial("logit")))
  )

growth_estimates

```
```{r}
## extract daily growth advantage
rho_estimates <- growth_estimates %>%
  filter(term == "t") %>%
  select(country, estimate, std.error) %>%
  mutate(
    rho = estimate / 14,
    rho_se = std.error / 14,
    rho_lo = rho - 1.96*rho_se,
    rho_hi = rho + 1.96*rho_se
  )

rho_estimates
```

```{r}
## run model with country-specific parameters based on proportion of XBB1.5 on a specific month
# start_month <- lubridate::ym("2022-12")
# 
# ## extract country-specific proportions on start date
# xbb <- xbb %>%
#   mutate(
#     ym = lubridate::ym(str_sub(as.character(date), 1, 7))
#   )
#          
# start_df <- xbb %>%
#   filter(ym == start_month) %>%
#   filter(date == min(date)) %>%
#   arrange(country)

## country-specific start dates (first week when variant detected)
start_dates <- xbb %>%
  group_by(country) %>%
  summarise(start_date = min(date)) %>%
  select(country, start_date)

## country-specific growth advantage estimates (based on first 10 weeks)
start_rho <- rho_estimates %>%
  select(country, rho, rho_se)

## country-specific starting proportions (based on first two-week period when variant first detected)
start_props <- xbb %>%
  filter(t == 1) %>%
  select(country, xbb_prop)

## country-specific starting proportions (mean of first two readings)
# start_props <- xbb %>%
#   filter(t <= 2) %>%
#   group_by(country) %>%
#   summarise(
#     xbb = sum(xbb_sequences),
#     total = sum(total_sequences)) %>%
#   mutate(
#     xbb_prop = xbb / total
#   ) %>%
#   select(country, xbb_prop)

start_df <- start_dates %>%
  left_join(start_rho) %>%
  left_join(start_props)
```

```{r}
countries <- start_df %>% select(country) %>% pull() # vector of country names
start_props <- start_df %>% select(xbb_prop) %>% pull() # vector of country-specific variant proportions
start_rho <- start_df %>% select(rho) %>% pull()
start_dates <- start_df %>% select(start_date) %>% pull() # start date of analysis
t_end <- 120 # length of estimation period in days
#rho <- 0.08 # assumed growth advantage for new variant at the start of the analysis period
#rho_min <- rho * 0.75 # if growth advantage is 25% lower or 50% higher
#rho_max <- rho * 1.25

estimate_pt_country <- function(input_countries, p, rho, start) { # arguments are country, starting proportion, growth advantage, start date
  
  mu <- (1 - p) / p
  rho_min <- rho * 0.75 # if growth advantage is 25% lower or 50% higher
  rho_max <- rho * 1.25
  
  ## calculate date when proportion reaches 50%
  t50 <- -1 * rho^-1 * (log(1-0.5) - log(mu * 0.5))
  
  d <- tibble(
    country = rep(input_countries, t_end+1),
    t = c(0:t_end),
    date = start + t
  )
  
  d <- d %>%
    mutate(
      pt = 1 / (1 + mu * exp(1)^-(rho * t)),
      pt_min = 1 / (1 + mu * exp(1)^-(rho_min * t)),
      pt_max = 1 / (1 + mu * exp(1)^-(rho_max * t)),
      t50 = start + t50
    )

}

l <- list(input_countries = countries, p = start_props, rho = start_rho, start = start_dates) # match inputs to function arguments
country_list <- pmap(l, estimate_pt_country) # generate estimates
xbb_predict <- country_list %>% bind_rows()

## merge with observed data
xbb_predict <- xbb_predict %>%
  left_join(xbb, by = c("country", "date"))

## extract mid-points for plotting
xbb_midpoint <- xbb_predict %>%
  group_by(country) %>%
  summarise(t50 = min(t50))
```

```{r, fig.width=16, fig.height=8}
## plot
xbb_predict %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = pt)) +
  geom_ribbon(aes(ymin = pt_min, ymax = pt_max), fill = "gray80", alpha = 0.5) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), size = 0.5, col = "gray50") +
  geom_point(aes(y = xbb_prop)) +
  geom_hline(aes(yintercept = 0.5), col = "black", linetype = "dotted") +
  geom_vline(data = xbb_midpoint, aes(xintercept = t50), col = "black", linetype = "dotted") +
  facet_wrap(~country) +
  scale_x_date(date_labels = "%b-%y") +
  labs(
    x = "",
    y = "Proportion XBB.1.5",
    caption = "Growth advantage is estimated using first 10 weeks of variant detection\nShaded area represents uncertainty bounds based on growth advantage being 25% lower or higher"
  ) 
```
```{r}
## models with growth advantage estimated using all data
growth_estimates2 <- xbb %>%
  group_by(country) %>%
  group_modify(
    # Use `tidy`, `glance` or `augment` to extract different information from the fitted models.
    .f = ~ broom::tidy(glm(xbb_prop ~ t, data = ., weights = total_sequences, family = binomial("logit")))
  )

## predicted values
predict_df <- xbb %>%
  group_modify(
    .f = ~ broom::augment(glm(xbb_prop ~ t, data = ., weights = total_sequences, family = binomial("logit")), se_fit = TRUE)
  ) %>%
  mutate(
    fitted_prop = boot::inv.logit(.fitted),
    fitted_prop_lo = boot::inv.logit(.fitted - 1.96 * .se.fit),
    fitted_prop_hi = boot::inv.logit(.fitted + 1.96 * .se.fit),
    t = t - 1
  )

## merge dates
predict_df <- predict_df %>%
  left_join(start_df %>% select(country, start_date)) %>%
  mutate(
    date = start_date + (t * 7)
    )

growth_estimates2
```

```{r}
## extract daily growth advantage
rho_estimates <- growth_estimates2 %>%
  filter(term == "t") %>%
  select(country, estimate, std.error) %>%
  mutate(
    rho = estimate / 14,
    rho_se = std.error / 14,
    rho_lo = rho - 1.96*rho_se,
    rho_hi = rho + 1.96*rho_se
  )

rho_estimates

```

```{r, fig.width=16, fig.height=8}
## plot observed vs fitted
predict_df %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = fitted_prop)) +
  geom_ribbon(aes(ymin = fitted_prop_lo, ymax = fitted_prop_hi), fill = "gray80", alpha = 0.5) +
  #geom_errorbar(aes(ymin = Lower, ymax = Upper), size = 0.5, col = "gray50") +
  geom_point(aes(y = xbb_prop)) +
  geom_hline(aes(yintercept = 0.5), col = "black", linetype = "dotted") +
  geom_vline(data = xbb_midpoint, aes(xintercept = t50), col = "black", linetype = "dotted") +
  facet_wrap(~country) +
  scale_x_date(date_labels = "%b-%y") +
  labs(
    x = "",
    y = "Proportion XBB.1.5",
    caption = "Growth advantage is estimated using first 10 weeks of variant detection\nShaded area represents uncertainty bounds based on growth advantage being 25% lower or higher"
  ) 
```
```{r}
## import vaccination data from OWID
vax_data <- read_csv(here("data", "owid-covid-data.csv"))
#uk_countries <- c("United Kingdom", "England", "Scotland", "Wales", "Northern Ireland")

vax_data2 <- vax_data %>%
  filter(location %in% c(countries, "England")) %>%
  select(country = location, date, people_vaccinated_per_hundred, people_fully_vaccinated_per_hundred, total_vaccinations_per_hundred, total_boosters_per_hundred) %>%
  filter(country != "United Kingdom") %>%
  mutate(
    country = if_else(country == "England", "United Kingdom", country) # use England as indicator of vaccine coverage across UK
  ) %>%
  left_join(
    start_df %>% select(country, start_date)
  ) 

## use highest vaccine uptake figures for 2022
vax_data2 <- vax_data2 %>%
  mutate(
    year = lubridate::year(date)
  ) %>%
  group_by(country, year) %>%
  summarise(
    vaccinated_2022 = max(people_vaccinated_per_hundred, na.rm = TRUE),
    fully_vaccinated_2022 = max(people_fully_vaccinated_per_hundred, na.rm = TRUE),
    total_vaccinations_2022 = max(total_vaccinations_per_hundred, na.rm = TRUE),
    total_boosters_2022 = max(total_boosters_per_hundred, na.rm = TRUE)
  ) %>%
  ungroup %>%
  filter(year == 2022)

## reshape to long format
vax_data2_long <- vax_data2 %>%
  pivot_longer(-c(country, year), names_to = "indicator", values_to = "value")

## merge with growth estimates data
vax_data2_long <- vax_data2_long %>%
  left_join(rho_estimates, by = "country")
```

```{r, fig.height=9, fig.width = 12}
## plot vaccine coverage indicators against daily growth rate for different countries
vax_data2_long %>%
  ggplot(aes(x = value, y = rho)) +
  geom_point() +
  geom_text(aes(label = country), size = 2, vjust = 1, hjust = 1) +
  facet_wrap(~indicator, scales = "free_x") + 
  scale_y_continuous(limits = c(0.04,0.1), breaks = seq(0.04, 0.1, 0.02))
```


```{r}
## Canada province/territory-specific data
## from: https://virusseq-dataportal.ca/explorer
## data downloaded on 2023-04-04 for the period from 2023-10-01 onwards
ca_variants_by_pt <- read_tsv(here("data", "virusseq-metadata-export-2023-04-04T17_05_45.405Z.tsv"))

## count XBB1.5 and total sequences by PT and sample collection date
xbb_by_pt <- ca_variants_by_pt %>%
  mutate(
    xbb15 = if_else(`lineage name` == "XBB.1.5", 1, 0, missing = 0)
  ) %>%
  group_by(study_id, `sample collection date`) %>%
  summarise(xbb15 = sum(xbb15), total = n())

## get week start date for sample date
xbb_by_pt <- xbb_by_pt %>%
  mutate(
    sample_week = lubridate::floor_date(`sample collection date`, unit = "week", week_start = 1)
  )

## collapse by sample week
xbb_by_pt <- xbb_by_pt %>%
  group_by(study_id, sample_week) %>%
  summarise(
    xbb15 = sum(xbb15),
    total = sum(total)
  ) %>%
  mutate(
    xbb_prop = xbb15 / total
  )
```
```{r}
## plot variant proportion by PT
xbb_by_pt %>%
  ggplot(aes(x = sample_week, y = xbb_prop)) +
  geom_col() +
  facet_wrap(~study_id)
```
```{r}
## find earliest detection week by province
xbb_by_pt <- xbb_by_pt %>%
  filter(xbb_prop > 0) %>%
  arrange(study_id, sample_week) %>%
  group_by(study_id) %>%
  mutate(
    earliest_date = min(sample_week)
    ) %>%
  ungroup

## set start proportions as average of first two detection weeks
start_props <- xbb_by_pt %>%
  group_by(study_id) %>%
  filter(row_number() <=2) %>%
  summarise(
    xbb15 = sum(xbb15),
    total = sum(total),
    xbb_prop = xbb15 / total
  )

```

